### 2019-07-04 manuscript.tex
Update manuscript.tex

Laurentia refers to the craton that forms the Precambrian core of North America. Laurentia is comprised of multiple Archean cratons that had unique histories prior to their amalgmation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly (Hoffman, 1988). Collision between the Superior Craton and the composite Slave+Rae+Hearne+Nain cratons that resulted in the the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Corrigan et al., 2009). Terminal collison recorded in the Trans-Hudson orogeny is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims (Skipton et al., 2016; Weller and St-Onge, 2017). A period of accretionary and collision orogenesis is recorded in the constituent cratons and terranes of Laurentia leading up to the terminal collison of the Trans-Hudson orogeny. This story of rapid Paleoproterozoic amalgamation of  Laurentia's constituent cratons, including the terminal Trans-Hudson orogeny, was synthesized in the seminal paper \textit{United Plates of America} (Hoffman, 1988) and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of the more major cratonic blocks: the Thelon orogeny associated with the collision between the Slave craton and the Rae craton ca. 2.0 to 1.9 Ga (Hoffman, 1988); the Snowbird orogeny associated with ca. 1.89 Ga collision between the Rae and Hearne cratons and associated terranes (Berman et al., 2007); the XXXXXX orogeny associated with ca. 1.86 Ga collision between the Rae and Nain cratons (St-Onge et al., 2009). As for the Wyoming craton, many models posit that it was conjoined with Hearne and associated cratons as the time of the Trans-Hudson orogeny (e.g. St-Onge et al., 2009; Pehrsson et al., 2015) although a contrasting view has been been proposed that it arrived ca. 1.72 Ga after the accretionary Yavapai and Penokean orogenies (Kilian et al. 2016).

In the paleogeographic model framework of Pehrsson et al. (2015), the cratonic and terrane collisions leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior craton from the composite Slave+Rae+Hearne+Nain cratons (often referred to as the Churchill domain or plate; e.g. Skipton et al., 2016; Weller and St-Onge, 2017). This model posits that this period terminal collision is not only associated with the amalgmation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that including other major Paleoproterozoic cratons including Siberia, Congo/Sao Francisco, West Africa, and Amazonia (Pehrsson et al., 2015). Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a change in the style of Laurentia's growth as subsequent growth proceeding dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean cratons (\citealp{Whitmeyer2007a}; \ref{fig:Laurentia_map}). Major growth of Laurentia following the amalgmation of these Archean cratons occured associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny whose accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stablized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65–1.60 Ga Mazatzal Orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic. Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion. An interval of major plutonism occured ca. 1.48–1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces. In the latest Mesoproterozoic ca. 1.11-1.08 Ga, a major intracontinental rift co-located with the a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior craton and Paleoproterozoic provinces. This Midcontinent Rfit lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior. Major collisional orogenesis of the Grenvillian orogeny occured following Midcontinent Rift development

%The perphiral Penokean orogeny on the southern margin of the Superior craton coincided with the Trans-Hudson orogeny with the last evidence of the orogeny being undeformed plutons such as the ca. 1.78 Ga East Central Minnesota Batholith (Holm et al., 2005). The arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny

\begin{figure}
\centering
\includegraphics[width=6.5 in]{Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

This chapter utilizes the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications. The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them with grades meant to convey the confidence that the community has in these results. While many factors associated with paleomagnetic poles can be assessed quantitatively such as the Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occuring in Fall 2017 in Leirubakki, Iceland. An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are included and referred to as not-rated (`NR').

>>>>

Laurentia refers to the craton that forms the Precambrian core of North America. Laurentia is comprised of multiple Archean cratons that had unique histories prior to their amalgmation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1988a, Whitmeyer2007a}. Collision between the Superior Craton and the composite Slave+Rae+Hearne+Nain cratons that resulted in the the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collison recorded in the Trans-Hudson orogeny is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collision orogenesis is recorded in the constituent cratons and terranes of Laurentia leading up to the terminal collison of the Trans-Hudson orogeny. Coeval with the Trans-Hudson orogeny was the perphiral Penokean orogeny on the southern margin of the Superior cratron with the last evidence of that orogeny being ca. 1.78 undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent cratons, including the terminal Trans-Hudson orogeny, was synthesized in the seminal paper \textit{United Plates of America} \citep{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of the more major cratonic blocks: the Thelon orogeny associated with the collision between the Slave craton and the Rae craton ca. 2.0 to 1.9 Ga (Hoffman, 1988); the Snowbird orogeny associated with ca. 1.89 Ga collision between the Rae and Hearne cratons and associated terranes (Berman et al., 2007); the XXXXXX orogeny associated with ca. 1.86 Ga collision between the Rae and Nain cratons (St-Onge et al., 2009). As for the Wyoming craton, many models posit that it was conjoined with Hearne and associated cratons as the time of the Trans-Hudson orogeny (e.g. St-Onge et al., 2009; Pehrsson et al., 2015) although a contrasting view has been been proposed that it arrived ca. 1.72 Ga after the accretionary Yavapai and Penokean orogenies (Kilian et al. 2016).

In the paleogeographic model framework of \cite{Pehrsson2015a}, the cratonic and terrane collisions leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previous been a large oceanic tract separating the Superior craton from the composite Slave+Rae+Hearne+Nain cratons (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}. This model posits that this period terminal collision is not only associated with the amalgmation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that including other major Paleoproterozoic cratons including Siberia, Congo/Sao Francisco, West Africa, and Amazonia \citep{Pehrsson2015a}. Following the Trans-Hudson orogeny, the locus of orogenesis migrated to the exterior of Laurentia. This change marks a change in the style of Laurentia's growth as subsequent growth proceeding dominantly through accretion of juvenile crust along the southern and eastern margin of the nucleus of Archean cratons (\citealp{Whitmeyer2007a}; \ref{fig:Laurentia_map}). Major growth of Laurentia following the amalgmation of these Archean cratons occured associated with the arc-continent collision of the ca. 1.71 to 1.68 Ga Yavapai orogeny whose accretion was followed by widespread emplacement of granitoid intrusions \citep{Whitmeyer2007a}. These intrusions are hypothesized to have stablized the juvenile accreted terranes that subsequently remained part of Laurentia \citep{Whitmeyer2007a}. Subsequent accretionary orogenesis of the ca. 1.65–1.60 Ga Mazatzal Orogeny and associated plutonism lead to further crustal growth in the latest Paleoproterozoic. Laurentia's growth continued in the Mesoproterozoic along the southeast margin through further juvenile terrane and arc accretion. An interval of major plutonism occured ca. 1.48–1.35 Ga leading to the formation of A-type granitoids throughout both Mesoproterozoic and Paleoproterozoic provinces. In the latest Mesoproterozoic ca. 1.11-1.08 Ga, a major intracontinental rift co-located with the a large igneous province formed in Laurentia's interior leading to extension within the Archean Superior craton and Paleoproterozoic provinces. This Midcontinent Rfit lead to the formation of a thick succession of volcanics and mafic intrusions that are well-preserved in Laurentia's interior.  Midcontinent Rift development ceased as major collisional orogenesis of the Grenvillian orogeny began \citep{Swanson-Hysell2019a}. The Grenvillian orogeny was a protracted interval of continent-continent collision (ca. 1.09 to 0.98 Ga) which is interpreted to have resulted in the development of a thick orogenic plateau \citep{Rivers2008a} and is typically to interpretted to have been a major collisional event associated with the amalgamation of continents into the supercontinent Rodinia \citep{Hoffman1991a}.

\begin{figure}
\centering
\includegraphics[width=6.5 in]{Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean cratons (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \cite{Whitmeyer2007a}). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel while the diamonds are additional not-rated results from the Paleomagia database.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

This chapter utilizes the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications. The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results. While many factors associated with paleomagnetic poles can be assessed quantitatively such as the Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occuring in Fall 2017 in Leirubakki, Iceland. An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR').
